Improved Approximation Algorithms for the Maximum Carpool Matching Problem

### 2017-01-30 src/abstract.tex
Merge Dror

The \textsc{\CARPOOL{}} problem is a star packing problem in directed graphs.
Formally, given a directed graph $G = (V, A)$,
a capacity function $ c: V \rightarrow \N $,
and a weight function $w : A \rightarrow \R $,
a feasible \emph{carpool matching} is a triple 
$(P, D, M)$, where $P$ (passengers) and $D$ (drivers) form a partition of $V$, 
and $M$ is a subset of $A \cap (P \times D)$,
under the constraints that for every vertex $d \in D$, 
$\din(d) \leq c(d)$, 
and for every vertex $p \in P$, $\dout(p) \leq 1$.
In the \textsc{\CARPOOL{}} problem we seek for a matching $(P, D, M)$ that maximizes the
total weight of $M$.

The problem arises when designing an online carpool service, 
such as Zimride~\cite{zimride}, 
that tries to connect between passengers and drivers based on (arbitrary) similarity function.
The problem is known to be NP-hard, 
even for uniform weights and without capacity constraints.

we show that the problem can be formulated as
unconstarints submodular maximization problem, 
thus it has a 2-approximation algorithm.
For the unweighted variant of the problem when the maximum possible capacity, $C$, is
fixed, we give a $(\frac{C + 1}{2C} - \varepsilon)$-approximation algorithm
for the any $\varepsilon > 0$.

>>>>

The \carpool problem is a star packing problem in directed graphs.
Formally, given a directed graph $G = (V, A)$, a capacity function $c:
V \rightarrow \N$, and a weight function $w : A \rightarrow \R $, a
feasible \emph{carpool matching} is a subset of arcs, $M \subseteq A$,
such that every $v \in V$ satisfies:%
\begin{inparaenum}[(i)]
\item $\din[M](v) \cdot \dout[M](v) = 0$,
\item $\din[M](v) \leq c(v)$, and 
\item $\dout[M](v) \leq 1$.
\end{inparaenum}
A vertex $v$ for which $\dout[M](v) = 1$ is a \emph{passenger}, and a
vertex for which $\dout[M](v) = 0$ is a driver who has $\din[M](v)$
passengers.  In the \carpool problem the goal is to find a matching
$M$ of maximum total weight.

The problem arises when designing an online carpool service, such as
Zimride~\cite{zimride}, which tries to connect between passengers and
drivers based on (arbitrary) similarity function.  The problem is
known to be NP-hard, even for uniform weights and infinite capacities.
%without capacity constraints.

We show that the problem can be formulated as an unconstrained
submodular maximization problem, thus it admits a
$\frac{1}{2}$-approximation algorithm.  For the unweighted variant of
the problem when the maximum possible capacity, $C$, is fixed, we give
a $(\frac{C + 1}{2C} - \varepsilon)$-approximation algorithm for the
any $\varepsilon > 0$.

We also show that the
problem remains APX-hard even if $C \leq 3$.

We also consider an extension, called \gcp, in which each vertex
represents a group of passengers.  Group sizes are non-uniform, and we
consider the case where a group size may change as a function of the
driver.  Formally, each arc $(u,v) \in A$ has a size $s(u,v) \in \N$,
and the constraint $\din[M](v) \leq c(v)$ is replaced with
$\sum_{u:(u,v) \in M} s(u,v) \leq c(v)$.  We show that \gcp cannot be
formulated as an unconstrained submodular maximization problem.
Nevertheless, we give a $(\frac{1}{2} - \varepsilon)$-approximation
algorithm for this variant, and argue that the local search algorithm
can be generalized to this variant of the problem when $C$ is fixed.

### 2017-01-30 src/group.tex
Merge Dror

We now consider a variant of the problem when, in addition to the capacity
function $c:V \to \N$, and the weight (profit) function $w:A \to \R$, we also
have a size function $s:A \to \N$, and the constraint $\din[M] \leq c(v)$
is replaced with the constraint $\sum_{(u,v)}s(u,v) \leq c(v)$.
The motivation for this variant of the problem is when we want to allow a group
of people to travel together in the same car, thus $s$ is the size of the group.

We start by showing that this variant of the problem does not
fit the submodular maximization formulation anymore.
We show, however, that the problem still has a 
$\frac{1}{2} - \varepsilon$-approximation algorithm.
Finally, we argue that the local search algorithm can be trivially generalized
to handle the new variant of the problem while keeping the same approximation
guarantees.

@@

\begin{figure}
\caption{
\label{fig:not submodular}
Consider the following instance where node capacity are written inside nodes, 
arcs are labeled with their size and all weights are equal to 1.
It is easy to see that if $f(S)$ is the optimal solution where $S$ is the set of
drivers, then $f(A) + f(B) = 2 \leq 3 = f(A \cup B) + f(A \cap B)$,
thus, $f$ is not submodular. 
}
\begin{center}
\input{fig-not-submodular}
\end{center}
\end{figure}

>>>>

We now consider a variant of the problem when, in addition to the
capacity function $c:V \to \N$, and the weight (profit) function
$w:A \to \R$, we also have a size function $s:A \to \N$, and the
constraint $\din[M] \leq c(v)$ is replaced with the constraint
$\sum_{(u,v)}s(u,v) \leq c(v)$.  The motivation for this variant of
the problem is when we want to allow a group of people to travel
together in the same car, thus $s$ is the size of the group.  We start
by showing that this variant of the problem does not fit the
submodular maximization formulation anymore. We show, however, that
the problem still has a $(2 +\eps)$-approximation algorithm.  Finally,
we argue that the local search algorithm can be trivially generalized
to handle the new variant of the problem while keeping the same
approximation guarantees.

%%%%%

\subsection{Submodular Maximization}

@@

\begin{figure}
\begin{center}
\input{fig-not-submodular}
\end{center}
\caption{Consider the following instance where node capacity are
written inside nodes, arcs are labeled with their size and all weights
are equal to 1.  It is easy to see that if $f(S)$ is the optimal
solution where $S$ is the set of drivers, then $f(A) + f(B) = 2 \leq 3
= f(A \cup B) + f(A \cap B)$, thus, $f$ is not submodular.}
\label{fig:not submodular}
\end{figure}

%%%%%

\subsection{A $(2+\eps)$-approximation Algorithm}

\todo[inline]{write a theorem and a proof}

We now argue that this algorithm can be generalized to solve the group carpool
matching problem.
The main concern when trying to adopt the algorithm to the group carpool
matching is how to determine if a vertex can be improved.
Observe that Remark~\ref{rem:improve} does not hold anymore.
If the maximum possible capacity is bounded, one can test for improvement of
$v$ by considering all the subsets of edges intersecting $v$. 
When the capacity is unbounded, however, it is easy to see that efficient
way to test improvement implies a solution to the Knapsack problem.
Indeed, if this is the case, one can test for improvement by using a known FPTAS
for Knapsack, this is in the cost of additional $\varepsilon$ reduction in the
approximation ratio of the algorithm.

%%%%%

\subsection{A $(\frac{2\cmax}{\cmax+1} + \eps)$-approximation Algorithm}

\todo[inline]{complete}

We now argue that the local sear

### 2017-01-30 src/hardness.tex
Merge Dror

Formally, the (unweighted) minimum dominating set problem is defined as
$\min_{D}(|D|)$,
where $D$ is any feasible dominating set.
It is also easy to see that the (unweighted, undirected, uncapacitated) 
maximum carpool matching problem can be defined as $max_{D}(|V \setminus D|)$, 
where $D$ is any feasible dominating set.
The \textsc{Dominating Set-B} is the problem of finding minimum dominating set on 
graphs with degree bounded by $B$,
and it was shown to be APX-hard (for $B \geq 3$) by 
Papadimitriou et al.~\cite{papadimitriou1988optimization}.

Let $G$ be a b-bounded degree graph,
let $D^*(G)$ be a minimum dominating set in $G$,
and let $C^*(G)$ be a maximum carpool matching in $G$, 
then the following lemmas hold:

\begin{lemma}
$|C^*| = |V| - |D^*|$
\end{lemma}

\begin{proof}
The proof follows immediately from the definitions of the two problems. 
\end{proof}

\begin{lemma}
\label{lm:optc_leq_boptd}
$|C^*| \leq b \cdot |D^*|$
\end{lemma}

\begin{proof}
The proof follows directly from the observation that $|D^*| \geq \frac{|V|}{b}$.
\end{proof}

Let $C(G)$ be any feasible carpool matching on $G$, then the following lemma holds:

\begin{lemma}
\label{lm:erreq}
$|C^*| - |C| = (|V| - |C|) - |D^*|$
\end{lemma}

\begin{proof}
$$
|C^*| - |C| 				= 
|V| - |D^*| - |C| 	= 
(|V| - |C|) - |D^*|
$$
\end{proof}

Let $D(G)$ be any feasible dominating set in $G$, 
we argue that $f(G) = G$, and $g(C(G)) = V \setminus C(G)$ 
define a linear reduction (L-reduction) from the \textsc{Dominating Set-B} problem to the 
\textsc{\CARPOOL{}} problem.
\begin{proof}
By lemma~\ref{lm:optc_leq_boptd} we get that $C^*(f(G)) = C^*(G) \leq b \cdot D^*(G)$,
and by lemma~\ref{lm:erreq} we get that 
\begin{equation}
\begin{split}
|C^*(f(G))| - |C(G)|		& = |C^*(G)| - |C(G)| 			\\
							& = (|V| - |C(G)|) - |D^*(G)|	\\
							& = |g(C(G))| - |D^*(G)|
\end{split}
\end{equation}
\end{proof}

\begin{theorem}
The \textsc{\CARPOOL{}} problem is APX-hard.
\end{theorem}

\begin{proof}
The proof follows directly from the above linear reduction and the fact that 
\textsc{Dominating Set-B} is APX-hard. 
\end{proof}

>>>>

??????

As we mentioned in the introduction, the maximum spanning star forest
can not be approximated within a factor of $\frac{31}{32} + \epsilon$
for any $\epsilon > 0$, unless P$=$NP.  The proof, however, does not
hold for the case when the maximum capacity is constant.  In this
subsection we show that the problem remains APX-hard even in this
case.

??????

Formally, the (unweighted) \textsc{Minimum Dominating Set} problem is
defined as follows.  The input is an undirected graph $G = (V,E)$, and
a feasible solution, or a \emph{dominating set}, is a subset
$D \subseteq V$ that dominates $V$ namely such that
$D \cup \bigcup_{v \in D} N(v) = V$, where $N(v)$ is the neighborhood
of $v$.  The goal is to find a minimum cardinality dominating set.
%
\textsc{Minimum Dominating Set-$b$} is the special case of
\textsc{Minimum Dominating Set} in which
the maximum degree of a vertex in the input graph $G$ is bounded by
$b$.  The problem was shown to be APX-hard, for $b \geq 3$, by
Papadimitriou and Yannakakis~\cite{PapYan88}.

We now consider the unweighted and undirected special case of
the \carpool problem.  In this case, the input consists of an
undirected graph $G$ and a capacity function $c$, and the goal is to
find a carpool matching $M$ that maximizes $\abs{P_M}$.

Given an undirected graph $G$, let $D^*$ be a minimum cardinality
dominating set, and let $M^*$ be an optimal carpool matching with
respect to $G$ and the capacity function: $c(v) = \deg(v)$, for every
$v \in V$.

\begin{observation}
$\abs{P_{M^*}} + \abs{D^*} = \abs{V}$ 
\end{observation}
\begin{proof}
Given a carpool matching $M$, observe that $D_M \cup Z_M$ is a
dominating set.  In the other direction, a dominating set $D$ induces
a carpool matching of size $\abs{V \setminus D}$.
\end{proof}

We use this duality to obtain a hardness result for \carpool.

\begin{theorem}
The \carpool problem is APX-hard, even for undirected and unweighted
instances with maximum capacity bounded by $b$, for $b \geq 3$.
\end{theorem}
\begin{proof}
We prove the theorem by presenting an $L$-reduction from
\textsc{Minimum Dominating Set-$b$}.
(For details on $L$-reductions the reader is referred to~\cite{??}.)

We define a function $f$ from \textsc{Minimum Dominating Set-$b$}
instances to \carpool instances as follows: $f(G) = (G,c)$, where
$c(v) = \deg(v)$, for every $v \in V$.  Next, we define a function $g$
that given a carpool matching computes a dominating set as follows:
$g(M) = V \setminus P_{M}$.  Both $f$ and $g$ can be computed in
polynomial time.

Let $D^*$ be an optimal dominating set with respect to $G$, and let
$M^*$ be an optimal carpool matching with respect to $G$ and $c$.
Since $|D^*| \geq \frac{|V|}{b+1}$, it follows that 
\[
\abs{P_{M^*}} \leq b \abs{D^*}
~,
\]
In addition, if $M$ is a carpool matching, we have that
\[
\abs{D_M \cup Z_M} - \abs{D*}
= (\abs{V} - \abs{P_M}) - \abs{D^*}
= \abs{P_{M^*}} - \abs{P_M}
~.
\]
Hence, there is an $L$-reduction from \textsc{Minimum Dominating
Set-$b$} to unweighted and undirected \carpool with bounded capacity
$b$.
\end{proof}

### 2017-01-30 src/intro.tex
Merge Dror

Carpooling, is the sharing of car journeys so that more than one person travels
in a car.
Knapen et al.~\cite{knapen2013estimating} describe an automatic service
to match commuting trips.
Users of the service register their personal profile and a set of periodically
recurring trips, 
and the service advises registered candidates on how to combine their commuting
trips by carpooling.
The service acts in two phases.

@@

The second phase can be modeled in terms of graph theory.
Given a directed graph $G = (V, A)$.
Each vertex $v \in V$ corresponds to a user of the service and an arc
$(u, v)$ exists if the user corresponding to vertex $u$ is willing to
commute with the user corresponding to vertex $v$.
A capacity function $ c: V \rightarrow \N $ is defined
according to the number of passengers each user can drive if she was
selected as a driver.
A weight function $w : A \rightarrow \R $ defines the amount of
satisfaction $w(u, v)$,
that user $u$ gains when riding with user $v$.

A feasible \emph{carpool matching} (matching) is a triple 
$(P, D, M)$, where $P$ and $D$ form a partition of $V$, 
and $M$ is a subset of $A \cap (P \times D)$,
under the constraints that for every driver $d \in D$, 
$\din(d) \leq c(d)$, 
and for every passenger $p \in P$, ${\dout(p) \leq 1}$.
In the \textsc{\CARPOOL{}} problem we seek for a matching $(P, D, M)$ that maximizes the
total weight of $M$.
In other words, the \textsc{\CARPOOL{}} problem is about finding a set of 
(directed toward the center) vertex disjoint stars 
that maximizes the total weights on the arcs.
Figure~\ref{fig:carpool} is an example of the \textsc{\CARPOOL{}} problem.

@@

Hartman et al.~\cite{hartman2013optimal} proved that the
\emph{\CARPOOL{}} problem considered in this paper is NP-hard,
and that the problem remains NP-hard even for a binary weight function when
the capacity function $c(v) \leq 2$ for every vertex in $V$.
It is also worth mentioning, that in the undirected, uncapacitated, unweighted
variant of the problem, the set of drivers in an optimal solution
form a minimum dominating set.
When the set of drivers is known in advanced, however, the problem becomes
tractable and can be solved using a reduction to a flow network problem.

Agatz et al.~\cite{agatz2012optimization} outlined the optimization challenges
that arise when developing technology to support ride-sharing and survey the
related operations research models in the academic literature.  
Hartman et al.~\cite{hartman2014theory} designed several heuristic algorithms
for the \CARPOOL{} problem and compared 
their performance on real data.
Other heuristic algorithms were developed as well~\cite{knapen2014exploiting}.
Arkin et al.~\cite{arkin2004approximations}, considered other variants of
capacitated star packing where a capacity vector is given as part of the input and 
capacities need to be assigned to vertices.

Nguyen et al.~\cite{nguyen2008approximating} considered the \textsc{spanning star forest} problem
(the undirected, uncapacitated, unweighted variant of the problem).
They proved the following results:
\begin{enumerate*}
\item
there is a polynomial-time approximation scheme for planner graphs;
\item 
there is a polynomial-time $\frac{3}{5}$-approximation algorithm for graphs;
\item 
there is a polynomial-time $\frac{1}{2}$-approximation algorithm for weighted graphs.
\end{enumerate*}
They also showed how to apply the spanning star forest model to aligning multiple
genomic sequences over a tandem duplication region.
Chen et al.~\cite{chen2007improved} improved the approximation ratio to 0.71,
and also showed that the problem can not be approximated to within a factor of
$\frac{31}{32} + \epsilon$ for any $\epsilon > 0$ under the assumption 
that $\text{P} \neq \text{NP}$.
It is not clear, however, if any of the technique used to address the
\textsc{spanning star forest} problem can be generalized to approximate the
directed capacitated variant.

In section~\ref{sec:sub} we show that the problem can be formulated as
unconstarints submodular maximization problem, thus it has a 2-approximation
algorithm.
In section~\ref{sec:local} we consider the unweighted variant of the problem.
We show that when the maximum possible capacity, $C$, is fixed, there is a 
$(\frac{C + 1}{2C} - \varepsilon)$-approximation algorithm for the any $\varepsilon > 0$.

>>>>

Carpooling is the sharing of car journeys so that more than one
person travels in a car.  Knapen et al.~\cite{knapen2013estimating}
describe an automatic service to match commuting trips.  Users of the
service register their personal profile and a set of periodically
recurring trips, and the service advises registered candidates on how
to combine their commuting trips by carpooling.  The service acts in
two phases.

@@

The second phase can be modeled in terms of graph theory.  Given a
directed graph $G = (V, A)$, each vertex $v \in V$ corresponds to a
user of the service and an arc $(u, v)$ exists if the user
corresponding to vertex $u$ is willing to commute with the user
corresponding to vertex $v$.  A capacity function $ c:
V \rightarrow \N $ is defined according to the number of passengers
each user can drive if she was selected as a driver.  A weight
function $w : A \rightarrow \R $ defines the amount of satisfaction
$w(u, v)$ that user $u$ gains when riding with user $v$.
%
%Let $(G,c,w)$ be a carpool instance.
If $(u,v) \in A$ implies that $(v,u) \in A$ and $w(u,v) = w(v,u)$, we
call the instance \emph{undirected}.  If $w(v,u) = 1$, for every
$(v,u) \in A$, then we call the instance
\emph{unweighted}.  If $c(v) = \infty$, for every $v$, then we
call the instance \emph{uncapacitated}.

Given a directed graph $G$ and a subset $M \subseteq A$, define
$\din[M](v) \eqdf \abs{\set{u :(u,v) \in M}}$ and
$\dout[M](v) \eqdf \abs{\set{u :(v,u) \in M}}$.
%
A feasible \emph{carpool matching} is a subset of arcs, $M \subseteq
A$, such that every $v \in V$ satisfies:%
\begin{inparaenum}[(i)]
\item $\din[M](v) \cdot \dout[M](v) = 0$,
\item $\din[M](v) \leq c(v)$, and 
\item $\dout[M](v) \leq 1$.
\end{inparaenum}
A vertex $v$ for which $\dout[M](v) = 1$ is a \emph{passenger}, and a
vertex for which $\dout[M](v) = 0$ is a driver who has $\din[M](v)$
passengers.  In the \carpool problem the goal is to find a matching
$M$ of maximum total weight, namely to minimize
$w(M) \eqdf \sum_{(v,u) \in M} w(v,u)$.  In other words, the \carpool
problem is about finding a set of (directed toward the center) vertex
disjoint stars that maximizes the total weight of the arcs.
Figure~\ref{fig:carpool} is an example of the \carpool problem.

@@

In this paper we also consider an extension of \carpool, called \gcp,
in which each vertex represents a group of passengers.  Group sizes
are non-uniform, and we consider the case where a group size may
change as a function of the driver.  Formally, each arc $(u,v) \in A$
is associated with a size $s(u,v) \in \N$, and the constraint
$\din[M](v) \leq c(v)$ is replaced with the constraint
$\sum_{u:(u,v) \in M} s(u,v) \leq c(v)$.
%
Notice that \textsc{Knapsack} is the special case, where only arcs
directed at a single vertex have non-zero weights.

%%%%%

\subparagraph{Related work.}
%
Hartman et al.~\cite{hartman2013optimal} proved that the \carpool
problem considered in this paper is NP-hard, and that the problem
remains NP-hard even for a binary weight function when the capacity
function $c(v) \leq 2$ for every vertex in $V$.  It is also worth
mentioning, that in the undirected, uncapacitated, unweighted variant
of the problem, the set of drivers in an optimal solution form a
minimum dominating set (see Section~\ref{sec:hardness}).  When the set
of drivers is known in advanced, however, the problem becomes
tractable and can be solved using a reduction to a flow network
problem.

Agatz et al.~\cite{agatz2012optimization} outlined the optimization
challenges that arise when developing technology to support
ride-sharing and survey the related operations research models in the
academic literature.  Hartman et al.~\cite{hartman2014theory} designed
several heuristic algorithms for the \carpool problem and compared
their performance on real data.  Other heuristic algorithms were
developed as well~\cite{knapen2014exploiting}.  Arkin et
al.~\cite{arkin2004approximations} considered other variants of
capacitated star packing where a capacity vector is given as part of
the input and capacities need to be assigned to vertices.

Nguyen et al.~\cite{nguyen2008approximating} considered
the \textsc{spanning star forest} problem (the undirected,
uncapacitated, unweighted variant of the problem).  They proved the
following results:
\begin{inparaenum}[(i)]
\item there is a polynomial-time approximation scheme for planner graphs;
\item there is a polynomial-time $\frac{3}{5}$-approximation algorithm for graphs;
\item there is a polynomial-time $\frac{1}{2}$-approximation algorithm for weighted graphs.
\end{inparaenum}
They also showed how to apply the spanning star forest model to
aligning multiple genomic sequences over a tandem duplication region.
Chen et al.~\cite{chen2007improved} improved the approximation ratio
to 0.71, and also showed that the problem can not be approximated to
within a factor of $\frac{31}{32} + \epsilon$ for any $\epsilon > 0$
under the assumption that $\text{P} \neq \text{NP}$.  It is not clear,
however, if any of the technique used to address the
\textsc{spanning star forest} problem can be generalized to approximate the
directed capacitated variant.

Kutiel~\cite{Kutiel16}

\todo[inline]{add a paragraph about the paper from arXiv}

\todo[inline]{Mention Goerge's paper with Amotz (ESA)}

%%%%%

\subparagraph{Our contribution.}
%
In Section~\ref{sec:sub} we show that the problem can be formulated as
unconstrained submodular maximization problem, thus it has a
randomized $2$-approximation algorithm due to~\cite{BFNS15}.  We also
provide a deterministic $(2+\eps)$-approximation algorithm, for any
$\eps>0$, which is based on local search.

In Section~\ref{sec:local} we consider the unweighted variant of the
problem.  We provide a $(\frac{2\cmax}{\cmax+1} + \eps)$-approximation
algorithm, for any $\eps>0$, where $\cmax$ is the maximum capacity.
We also show that \carpool is APX-hard even if $\cmax \leq 3$.
%
In Section~\ref{sec:group} we consider \gcp.  We first show that \gcp
cannot be formulated as an unonstrained submodular maximization
problem as \carpool.  We show, however, that this problem still admits
a $(2 + \eps)$-approximation algorithm, and argue that the local
search algorithm generalizes to \gcp as well.
